---
title: "NOAA Storm Events Database Analysis"
date: "Monday, April 20, 2015"
output: 
  html_document:
    keep_md: true
---

## An analysis of weather events and their impact on human health and economics

## Synopsis
This document represents an analysis of the National Oceanic and Atmospheric
Administration (NOAA) Storm Events Database. As described on the
Storm Events Database homepage[1], this database documents information from
storms or significant weather phenommena resulting in human casualties (i.e.
injuries or death), property damage, disruption to commerce, media attention
and/or significant meteorological events (e.g. record temperatures or
precipitation).  This analysis focuses on the weather events that are most
harmful with respect to human health and result in the greatest economic
consequences.

## Data Processing

#### Load Required Libraries
```{r}
library(ggplot2)
library(grid)
library(dplyr)
```

#### Initial Setup
```{r}
if(!(file.exists("./data") & file.info("./data")$isdir)) dir.create("./data")
download.file(
    paste0("https://d396qusza40orc.cloudfront.net/"
           ,"repdata%2Fdata%2FStormData.csv.bz2"
    )
    ,destfile = "./data/repdata_data_FStormData.csv.bz2"
)
write(date(), file = "./data/date_downloaded.txt")
```
This code block creates a subdirectory for the data, downloads the data and
creates a file to document the time the data was acquired.

#### Loading and Preprocessing the Data
```{r}
data <- read.csv(bzfile("./data/repdata_data_FStormData.csv.bz2"))
data$date <- as.Date(data$BGN_DATE, "%m/%d/%Y")
data <- filter(data, date > "1996-01-01")
data$damage = 0
for(i in 1:nrow(data)){
    if(data[[i,"PROPDMGEXP"]] == "K"){
        data[[i,"damage"]] <- data[[i,"damage"]] + 
            (data[[i,"PROPDMG"]] * 1000)
    }
    else if(data[[i,"PROPDMGEXP"]] == "M"){
        data[[i,"damage"]] <- data[[i,"damage"]] + 
            (data[[i,"PROPDMG"]] * 1000000)
    }
    else if(data[[i,"PROPDMGEXP"]] == "B"){
        data[[i,"damage"]] <- data[[i,"damage"]] + 
            (data[[i,"PROPDMG"]] * 1000000000)
    }
    if(data[[i,"CROPDMGEXP"]] == "K"){
        data[[i,"damage"]] <- data[[i,"damage"]] + 
            (data[[i,"CROPDMG"]] * 1000)
    }
    else if(data[[i,"CROPDMGEXP"]] == "M"){
        data[[i,"damage"]] <- data[[i,"damage"]] + 
            (data[[i,"CROPDMG"]] * 1000000)
    }
    else if(data[[i,"CROPDMGEXP"]] == "B"){
        data[[i,"damage"]] <- data[[i,"damage"]] + 
            (data[[i,"CROPDMG"]] * 1000000000)
    }
}
data <- mutate(data, casualties = FATALITIES + INJURIES)
```
This code block loads the data and executes several preprocessing steps to
facilitate subsequent analysis.  Preprocessing steps include:  

- Creating a variable (*date*) by coercing the *BGN_DATE* column to the Date
class.

- Excluding data prior to 1996 via the filter function.  This is motivated by
information conveyed on the Storm Event Database details page[2] which states
that data prior to 1996 only contain a subset of the 48 event types that have
been documented in the subsequent data (1996 to current).  By excluding data
prior to 1996 the analysis ensures equal consideration for all event types.

- Creating a singe variable (*damage*) to quantify economic impact resulting
from damage or loss to private property, public infrastructure and crops.
Damage estimates are represented within the data in the variables *PROPDMG*
(Property Damage), *PROPDMGEXP* (Property Damage Exponent), *CROPDMG* (Crop
Damage) and *CROPDMGEXP* (Crop Damage Exponent).  Per Section 2.7 of the NWS
Directive 10-1605[3],

> Estimates should be rounded to three significant digits, followed by an
alphabetical character signifying the magnitude of the number, i.e., 1.55B for
$1,550,000,000. Alphabetical characters used to signify magnitude include “K”
for thousands, “M” for millions, and “B” for billions. If additional precision
is available, it may be provided in the narrative part of the entry.  

- The *PROPDMG* and *CROPDMG* variables are of class numeric and do not contain
NAs, generally following the "three significant digits" guidance as cited above.
However, the *PROPDMGEXP* and *CROPDMGEXP* variables contain some observations
outside of the expected values "K", "M" and "B."  For this analysis the *damage*
variable represents the sum of the values expressed in the aforementioned
variables for each observation where exponents are in the expected values set.
Thus, each observation of *damage* will be the sum of (*PROPDMG* * 
*PROPDMGEXP*) + (*CROPDMG* * *CROPDMGEXP*) for each observation where
*PROPDMGEXP* or *CROPDMGEXP* is "K", "M" or "B."  Exponent values with
unexpected values and any additional information contained in *REMARKS* are
ignored.

- Creating a variable (*casualties*) to sum values from the *FATALITIES* and
*INJURIES* variables.  FUrther information about fatalities and injuries is
located in Section 2.6 of the NWS Directive 10-1605[3].

## Results

## Reference
[1] http://www.ncdc.noaa.gov/stormevents/  
[2] http://www.ncdc.noaa.gov/stormevents/details.jsp  
[3] http://www.ncdc.noaa.gov/stormevents/pd01016005curr.pdf  
